Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-08-01 README.Rmd
Removed site labels from CA and PCA plots. Some additional plots and tables exploring the components of the PCA
Final edits

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# dialpastrepository

@@

This repository contains the data and code for our paper:

> Authors, (YYYY). *`r Title`*. Name of journal/book
> <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed
> `r format(Sys.Date(), "%d %b %Y")`. Online at
> <https://doi.org/xxx/xxx>

@@

> Authors, (`r format(Sys.Date(), "%Y")`). *Compendium of R code and
> data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`.
> Online at <https://doi.org/xxx/xxx>

@@

-   [:file_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
-   [:file_folder: data](/analysis/data): Data used in the analysis.
-   [:file_folder: figures](/analysis/figures): Plots and other
    illustrations
-   [:file_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents prepared
    and collected during the analysis.

>>>>

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Exploring the composition of lithic assemblages in Mesolithic
south-eastern Norway"
```

# exploring-assemblages-se-norway

@@

<!-- This repository contains the data and code for the paper: -->

<!-- > Authors, (YYYY). *`r Title`*. Name of journal/book -->
<!-- > <https://doi.org/xxx/xxx> -->

Pre-print is online here:

>  Roalkvam, Isak (2021). *`r Title`*. Accessed
> `r format(Sys.Date(), "%d %b %Y")`. Online at
> <https://doi.org/xxx/xxx>

@@

> Roalkvam, Isak (`r format(Sys.Date(), "%Y")`). *Compendium of R code and
> data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`.
> Online at <https://doi.org/xxx/xxx>

@@

-   [:file_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.pdf`, suitable for reading (the code is replaced by figures
    and tables in this file)
-   [:file_folder: data](/analysis/data): Data used in the analysis.
-   [:file_folder: figures](/analysis/figures): Plots and other
    illustrations
-   [:file_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents prepared
    and collected during the analysis.

### 2021-08-01 analysis/paper/paper.Rmd
Final edits

```{r libraries, eval = TRUE, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

@@

(overview + sitemap) /
(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 13, heading.align = "left")
```

@@

lithics2 <- sites %>% 
  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
                           fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("no")

@@

wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))

@@

# Correlation with mean site date
secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
denscor <- cor(lithics2$mean_date, lithics2$lithic_dens)

secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
denscorflint <- cor(wabiflint$mean_date, wabiflint$flint_dens)

@@

```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns. Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
# fig.width = 10

# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

@@

```{r supp1, cache = FALSE}
# Supplementary table of radiocarbon dates
radiocarbon %>% 
  select(-excavation_project) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    material = "Material",
    lab_code = "Lab code",
    c14_bp = "C14-age",
    error = "Error"
  )
```

>>>>

```{r libraries, eval = TRUE, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(knitr)
library(kableExtra)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

# Table 1. Overfiew of chronological framework.
Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
            'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9200–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
framework <- c(rep("Glørstad (2010)", 4), rep("Reitan (2016)", 5))

df <- data.frame(Period, Duration, framework)

kable(select(df, -framework), booktabs = TRUE, 
      caption = "Chronological framework. Glørstad's (2010) divison of phases reflects the more established framework, to which Reitan (2016) has recently suggested considerable changes.", 
      col.names = NULL) %>% 
pack_rows(index = c("Glørstad (2004)" = 4, "Reitan (2016)" = 5))

@@

# Call to plot (using patchwork)
(overview + sitemap) /
(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

@@

```{r tab2}
# Table 2. Names of analysed sites, dating method used, and reported site age.
sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  kable(caption = "Analysed sites.", col.names = c("no",
                    "Site name",
                    "Dating method",
                    "Reported start (BCE)",
                    "Reported end (BCE)"),
        booktabs = TRUE, linesep = c(""))
```

@@

lithics2 <- sites %>% 
  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is to 
  # retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
            fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("no")

@@

# VDL and RFSL for all lithics
wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))

@@

# Correlation with mean site date, all lithics
secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
denscor <- cor(lithics2$mean_date, lithics2$lithic_dens)

# Correlation with mean site date, flint
secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
denscorflint <- cor(wabiflint$mean_date, wabiflint$flint_dens)

@@

```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns. Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used.
lithicspca <- lithicstr %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

@@

```{r supp1, cache = FALSE}
# Supplementary table of radiocarbon dates

select(radiocarbon, -excavation_project) %>% 
  kable(booktabs = TRUE, longtable = TRUE,
      col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
      linesep = c(""))
```
